Traditional agroforestry systems in Timor-Leste can store large amounts of carbon in both soil and biomass

### 2024-03-27 manuscript.Rmd
update with Marguerite's info on methodology

---
title: "Traditional agroforestry systems in Timor-Leste can store large amounts of carbon in both soil and biomass"
author: "Marguerite Cogné, Régis Peltier, Vincent Freycon, Alexis Toumazeau, Camille Piponiot"
date: "2024-03-19"
output:
  officedown::rdocx_document:
    number_sections: true
bibliography: references.bib
---

```{r setup, include=FALSE}
packages_needed = c(
  "ggplot2",      # version 3.4.4
  "ggrepel",      # version 0.9.4
  "ggspatial",    # version 1.1.9
  "ggpubr",       # version 0.6.0
  "readxl",       # version 1.4.3
  "data.table",   # version 1.14.8
  "knitr",        # version 1.45
  "BIOMASS",      # version 2.1.11
  "terra",        # version 1.7-55
  "car",          # version 3.1-2
  "lmerTest",     # version 3.1-3
  "performance",  # version 0.10.8
  "flextable"     # version 0.9.5
)

@@

Agroforestry has the potential to make agriculture more resilient while improving carbon sequestration by incorporating trees and other woody perennials into agricultural fields and diversifying agricultural landscapes. Traditional agricultural systems in tropical areas often include trees, but their carbon sequestration potential is not always well described, hindering their potential inclusion in climate change mitigation strategies. In this study, we quantified carbon storage in both vegetation biomass and soil in five traditional agroforestry systems (AFS) in Timor-Leste:  cropping systems with fallow, silvopastoral systems, young agroforests, home gardens and forest gardens. Our results show that these traditional AFS store large amounts of carbon: the AFS with the highest carbon stocks (forest gardens) stored an average of 210\ MgC\ ha$^{-1}$, similar to an old-growth tropical forest. Biomass carbon was strongly dependent on the type of AFS (which differed in tree cover), while soil carbon was less variable between AFS but more dependent on site. We found no relationship between the amount of carbon stored in biomass and soil. Our results highlight the high diversity of traditional agroforestry systems in Timor-Leste and their high capacity to store carbon.

@@

## Load plot general information
data_plot <- read_excel("data/DATA_BIOMASS_TL_SHORT.xlsx", "PLOT_GENERAL_INFO")

@@

  ## crop to reduce memory usage
  box = c(124.5, 127.5, -9.2, -8.3) # Timor Leste
  rst <- terra::rast(destfile)
  rst <- terra::crop(rst, terra::ext(box))
  writeRaster(rst, filename = destfile, overwrite = TRUE) 
}

@@

The research was carried out in the center of the Baucau district, on the north-eastern coast of Timor-Leste. Two communities, or suco (group of villages), were selected to represent two contrasting topo-geographical, climatic and historical cases: (i) Gariuai suco (long-lat coordinates: `r site_summary[site=="Gariuai", coord]`) is located on a plateau of Baucau limestone, with an annual rainfall of `r site_summary[site=="Gariuai", prec]` mm and an altitude ranging 300-750 m; and (ii) the village of Osso Luga in Samalari suco (long-lat coordinates: `r site_summary[site=="Osso Luga", coord]`) is located on the Matebian foothill based on Bobonaro Scaly Clay, with an annual rainfall of `r site_summary[site=="Osso Luga", prec]` mm and an altitude ranging `r site_summary[site=="Osso Luga", elev]` m (Figure\ \@ref(fig:map-sites)). The average soil properties differed between sites (Table\ S\@ref(tab:site-soil)), with Gariuai having Clay Loam soil texture compared to Osso Luga having clay soil texture. The two sites also have had different histories. Gariuai has been inhabited for at least 50 years, while Osso Luga was abandoned during the Indonesian occupation, and was only repopulated after 2000. During the period of abandonment, the various agroforestry systems in Osso Luga lay fallow.

@@

We used the typology described in @Cogne2024, developed in the same study area. This typology was based on the one proposed by @Nair1993;@Nair2021, and adapted to the local context with semi-structured interviews conducted in the same study area [@Cogne2024]. The five AFS identified were: crop system including a wooded fallow phase (CF), Sylvopastoral (SP), Young agroforest (YA), Home garden (HG) and Forest garden (FG). CF are plantations of a few non-perennial crops (such as maize, sweet potatoes, groundnuts and squash) which are left fallow after 1 to 3 years. SP land is a vast area (generally 200 to 500 ha) where cattle are grazed under a layer of trees. YA are agricultural plots where the farmer has been planting rows of trees, spaced at around 5x10m, for two to 10 years, usually with the help of projects. HG are plots of land close to houses, where crops and livestock (poultry, pigs, cattle, etc.) have been grown in the shade of numerous trees and palms for multiple uses (wood, fruit, fibre, etc.) for more than five years. FGs are generally former HGs (over 50 years old), which have been abandoned, for example as a result of the war, and are now used mainly for gathering with limited human influence.

@@

Taxonomic identification was carried out using tetum vernacular names, which were then linked to scientific names using iNaturalist and Pl\@ntnet application, Geoffrey Hull’s monograph [@Hull2006], the “Useful tropical plants” database [@Fern2014] as well as the Australian Northern Territory government's Native Plants Herbarium [@NT2013]. The correspondence between tetum names and scientific names is provided in the supplemental material (Table S\@ref(tab:tetum-names)).

@@

# stack tree tables
data_tot <- rbind(data_large[, col_keep], data_small[, col_keep])

# merge with plot information
data_tot <- merge(data_tot, data_plot, by = "N_PLOT")

# convert data_tot into data.table 
setDT(data_tot)

@@

# remove NA or <30 cm circumference values or < 100 cm outside subplots
data_tot <- subset(data_tot, 
                   !is.na(DIAM_130_CM) & 
                     DIAM_130_CM >= 30/pi &
                     (DIAM_130_CM >= 100/pi | grepl("Q", N_Q_ID)))

# split scientific names into genus and species
names <- tstrsplit(data_tot$SCIENTIFIC_NAME, " ")

# get genera
data_tot$GENUS <- names[[1]]

# get species
data_tot$SPECIES <- names[[2]]

# Timoneus should be Timonius
data_tot$GENUS[!is.na(data_tot$GENUS) & data_tot$GENUS == "Timoneus"] <- "Timonius"

# remove one tree with DBH = 0
data_tot <- subset(data_tot, DIAM_130_CM > 0)

@@

# get wood density
data_wd <- getWoodDensity(genus = data_tot$GENUS, species = data_tot$SPECIES)

# estimate mean AGC
data_tot$mean <- computeAGB(
  D = data_tot$DIAM_130_CM, 
  WD = data_wd$meanWD, 
  H = data_tot$HEIGHT
) * 0.4713 # conversion factor from biomass to carbon

@@

data_sim_agc <- data.table(
  data_tot,
  AGBmonteCarlo(
    D = data_tot$DIAM_130_CM, 
    WD = data_wd$meanWD, 
    errWD = data_wd$sdWD, 
    H = data_tot$HEIGHT, 
    errH = data_tot$HEIGHT*cvH, 
    Dpropag = "chave2004", 
    n = 1000,
    Carbon = TRUE
  )$AGC_simu
) |> 
  melt(
    measure.vars = c("mean", paste0("V", 1:1000)),
    variable.name = "iter",
    value.name =  "AGC"
  )

@@

## sum mean biomass in quadrats and entire plot
data_sim[, value := value + valueSmall]
```

@@

Soil organic carbon stocks (in Mg.ha$^{-1}$) were then calculated using the method recommended by @Poeplau2017 (M4), which correctly takes into account the coarse fraction of the soil:

@@

# add total carbon stocks
data_sim = rbind(data_sim, data_sim[, .(value = sum(value[variable != "SOC10"]), variable = "TOT"), .(N_PLOT, site, ID_AF, iter)])
# group total C stocks in plots 6 and 8 (otherwise will underestimate CF total C
# stocks); this will not be used for analysis, only to display average C stocks
# per AFS
data_sim <- rbind(
  data_sim[variable != "TOT" | !N_PLOT %in% c(6, 8)], 
  data_sim[variable == "TOT" & N_PLOT %in% c(6, 8), .(value = sum(value), N_PLOT = 6), .(site, ID_AF, variable, iter)]
)
```

@@

The estimated AGC and SOC stocks were log-transformed for all statistical tests, because we expect them to be log-normally distributed as they only take positive values. After testing for normality (Shapiro-Wilks test) and homoscedasticity (Levene’s test), we performed a two-way ANOVA on the log-transformed estimates of AGC and SOC stocks, grouped by AFS and site. To test the relationship between log-transformed SOC and AGC stocks, we fitted two linear mixed models. We predicted SOC stocks at two different depths (either 0-10 or 0-30 cm) as a function of AGC stocks (fixed effect) and site (Samalari or Osso-Luga; random effect).

@@

```{r group-comparison}
# shapiro wilk test - normality
dataC_plot[variable %in% c("AGC", "SOC"), .(
  shap = shapiro.test(log(mean))$p.value
), .(afs, variable)]
# levene test of homoscedasticity
dataC_plot[variable %in% c("AGC", "SOC"), 
           leveneTest(log(mean) ~ afs*site)$`Pr(>F)`[1], .(variable)]

# 2-way anova 
anova_table = 
  dataC_plot[variable %in% c("AGC", "SOC"), 
             .(pvalue = round(anova(lm(log(mean) ~ site + afs))$`Pr(>F)`[1:2], 3)), 
             .(variable)]
anova_table$factor = rep(c("site", "afs"), 2)
```

```{r lmereg}
data_reg = melt(dataC_plot, measure.vars = c("mean", "lwr", "upr"), variable.name = "stat") |> 
  dcast(plot+ site+ afs ~ variable + stat)

@@

Total carbon stocks varied between `r dataC_plot[variable=="TOT", round(min(mean))]` and `r dataC_plot[variable=="TOT", round(max(mean))]`\ MgC\ ha$^{-1}$, with an average of `r dataC_plot[variable=="TOT", round(mean(mean))]`\ MgC\ ha$^{-1}$ and a coefficient of variation of `r dataC_plot[variable=="TOT", round(sd(mean)/mean(mean)*100)]`%  (Table \@ref(tab:summary); Figure\ \@ref(fig:barplot)). The AFS with the lowest total carbon stocks was the `r dataC_afs[variable=="TOT", afs[which.min(mean)]]` with an average of `r dataC_afs[variable=="TOT", round(mean[which.min(mean)])]`\ MgC\ ha$^{-1}$; the AFS with the highest total carbon stocks was the `r dataC_afs[variable=="TOT", afs[which.max(mean)]]` with an average of `r dataC_afs[variable=="TOT", round(mean[which.max(mean)])]`\ MgC\ ha$^{-1}$ (Table\ \@ref(tab:summary); Figure\ \@ref(fig:barplot)).

@@

AGC stocks varied between `r dataC_plot[variable=="AGC", round(min(mean))]` and `r dataC_plot[variable=="AGC", round(max(mean))]`\ MgC\ ha$^{-1}$, with an average of `r dataC_plot[variable=="AGC", round(mean(mean))]`\ MgC\ ha$^{-1}$ and a coefficient of variation of `r dataC_plot[variable=="AGC", round(sd(mean)/mean(mean)*100)]`%  (Table \@ref(tab:summary); Figure\ \@ref(fig:barplot)). AGC stocks varied significantly between AFS (ANOVA p-value: `r subset(anova_table, variable=="AGC" & factor == "afs")$pvalue`, Table \@ref(tab:anova)), but not between site (ANOVA p-value = `r subset(anova_table, variable=="AGC" & factor == "site")$pvalue`, Table \@ref(tab:anova)). The AFS with the lowest AGC stocks was the `r dataC_afs[variable=="AGC", afs[which.min(mean)]]` with an average of `r dataC_afs[variable=="AGC", round(mean[which.min(mean)])]`\ MgC\ ha$^{-1}$; the AFS with the highest AGC stocks was the `r dataC_afs[variable=="AGC", afs[which.max(mean)]]` with an average of `r dataC_afs[variable=="AGC", round(mean[which.max(mean)])]`\ MgC\ ha$^{-1}$ (Table\ \@ref(tab:summary); Figure\ \@ref(fig:barplot)).

@@

SOC stocks varied between `r dataC_plot[variable=="SOC", round(min(mean))]` and `r dataC_plot[variable=="SOC", round(max(mean))]`\ MgC\ ha$^{-1}$, with an average of `r dataC_plot[variable=="SOC", round(mean(mean))]`\ MgC\ ha$^{-1}$ and a coefficient of variation of `r dataC_plot[variable=="SOC", round(sd(mean)/mean(mean)*100)]`%  (Table \@ref(tab:summary); Figure\ \@ref(fig:barplot)). No significant differences in SOC were observed between the five AFS studied (ANOVA p-value: `r subset(anova_table, variable=="SOC" & factor == "afs")$pvalue`, Table \@ref(tab:anova)). However, the measured SOC stocks differed between sites (ANOVA p-value: `r subset(anova_table, variable=="SOC" & factor == "site")$pvalue`, Table \@ref(tab:anova)). The SOC stocks in Gariuai were on average higher (`r dataC_site[site=="Gariuai" & variable=="SOC", round(mean)]`\ MgC\ ha$^{-1}$) than those in Osso Luga (`r dataC_site[site=="Osso Luga" & variable=="SOC", round(mean)]`\ MgC\ ha$^{-1}$; Figure\ \@ref(fig:barplot)).

@@

## Large C stocks in traditional AFS in Timor Leste

Compared to values reported in the literature for other AFS, the carbon stocks in this study were particularly high. The value of SOC stocks was similar to other reported values [@Feliciano2018], but some of the AFS studied here have particularly high stocks in their biomass [compared to the average of 74\ MgC\ ha$^{-1}$ in @Ma2020]. The AFS with the highest carbon stocks, forest gardens, had stocks close to those of old-growth tropical forests [270\ MgC\ ha$^{-1}$; @Lal2005]. These high carbon stocks could be explained by the fact that forest gardens have generally grown for more than 50 years, with little human intervention (Table S1). In terms of structure and carbon stocks, forest gardens should therefore resemble old secondary forests, which have been shown to recover more than 60% of the AGC stocks and almost all the SOC stocks of an old forest [@Poorter2021]. In addition, FGs are generally located close to villages, where soils are enriched by organic waste and animal manure, which can enhance tree growth [@Madejon2016].

@@

The location of AFS in the landscape is not random and could also explain the observed differences in AGC stocks. For example, most HGs are located close to houses, in alluvial valleys that are well supplied with water and where the soil is enriched with organic waste and animal manure, which is likely to increase soil fertility [@Palm2001]. The same applies to FGs, which are former HGs that have been more or less abandoned, and YAs, where the farmer has chosen the site with the intention of turning it into a HG. We therefore expect HGs, FGs and YAs to be located in areas that are more favorable for tree growth [@Wagner2012;@Madejon2016;@Treuer2018]. Conversely, CF and SP are mainly located on limestone plateaus where water is deep and soils are shallow, which could reduce tree growth and biomass [@Imada2008].

@@

## How to conserve and improve carbon stocks in traditional AFS in Timor Leste?

Although agroforestry can make agriculture more resilient while improving its carbon footprint and biodiversity conservation value [@Cardinael2021], poor adaptation to the local context can explain many failures to adopt new agroforestry techniques [@Coe2014]. Conversely, some traditional agroforestry techniques have evolved in a particular context (biophysical, social and economic) and are therefore adapted to it [@Gouyon1993;@Aumeeruddy1994;@Peltier1996]. This is the case of the traditional AFS studied here: their heritage and socio-economic values [as shown by @Cogne2024], in addition to their high carbon stock value, confirm the importance of their conservation, both nationally and internationally..

@@

@@

```{r tab.id = "summary", results = "show", tab.cap = "Mean and associated 95% confidence intervals (between parentheses) of total, aboveground carbon (AGC), belowground carbon (BGC), and soil organic carbon (SOC) stocks in each agroforestry system."}
# calculate median and 95% confidence interval at site level
dataC_afs[, value := paste0(
  signif(mean, 3), " (",
  signif(lwr_mean, 3), "-", 
  signif(upr_mean, 3), ")")] |> 
  subset(variable != "SOC10") |> 
  dcast(afs ~ variable, value.var = "value") |> 
  flextable(col_keys = c("afs", "TOT", "AGC", "BGC", "SOC"), 
            cwidth = c(0.9, 1.2, 1.3, 1.3, 1.2)) |> 
  compose(i = 1, j = "afs", part = "header", value = as_paragraph("Agroforestry system")) |>
  compose(i = 1, j = "TOT", part = "header", value = as_paragraph( "Total C stocks\n(MgC ha", as_sup("-1"), ")"))|> 
  compose(i = 1, j = "AGC", part = "header", value = as_paragraph( "AGC stocks\n(MgC ha", as_sup("-1"), ")"))|> 
  compose(i = 1, j = "BGC", part = "header", value = as_paragraph( "BGC stocks\n(MgC ha", as_sup("-1"), ")"))|> 
  compose(i = 1, j = "SOC", part = "header", value = as_paragraph( "SOC stocks\n(MgC ha", as_sup("-1"), ")"))
```

>>>>

---
title: "Traditional agroforestry systems in Timor-Leste can store large amounts of carbon in both soil and biomass"
author: "Marguerite Cogné, Régis Peltier, Vincent Freycon, Alexis Toumazeau, Camille Piponiot"
date: "2024-03-19"
output: 
  officedown::rdocx_document:
    reference_docx: word-styles-ref-02.docx
bibliography: references.bib
---

```{r setup, include=FALSE}
# Run with R version 4.3.2

packages_needed = c(
  "ggplot2",      # version 3.4.4
  "ggrepel",      # version 0.9.4
  "ggspatial",    # version 1.1.9
  "maps",         # version 3.4.2
  "ggpubr",       # version 0.6.0
  "readxl",       # version 1.4.3
  "data.table",   # version 1.14.8
  "knitr",        # version 1.45
  "BIOMASS",      # version 2.1.11
  "terra",        # version 1.7-55
  "car",          # version 3.1-2
  "lmerTest",     # version 3.1-3
  "lme4",         # version 1.1-35.1
  "Matrix",       # version 1.6
  "performance",  # version 0.10.8
  "flextable"     # version 0.9.5
)

@@

Agroforestry has the potential to make agriculture more resilient while improving carbon sequestration by incorporating trees and other woody perennials into agricultural fields and diversifying agricultural landscapes. Traditional agricultural systems in tropical areas often include trees, but their carbon sequestration potential is not always well described, hindering their potential inclusion in climate change mitigation strategies. In this study, we quantified carbon storage in both vegetation biomass and soil in five traditional agroforestry systems (AFS) in Timor-Leste:  cropping systems with fallow, silvopastoral systems, young agroforests, home gardens and forest gardens. Our results show that these traditional AFS store large amounts of carbon: the AFS with the highest carbon stocks (forest gardens) stored an average of 210\ Mg\ C\ ha$^{-1}$, similar to an old-growth tropical forest. Biomass carbon was strongly dependent on the type of AFS (which differed in tree cover), while soil carbon was less variable between AFS but more dependent on site. We found no relationship between the amount of carbon stored in biomass and soil. Our results highlight the high diversity of traditional agroforestry systems in Timor-Leste and their high capacity to store carbon.

@@

## Load plot general information
data_plot <- read_excel("data/DATA_BIOMASS_TL_SHORT.xlsx", "PLOT_GENERAL_INFO") |> 
  subset(N_PLOT != 8)  # remove plot 8: no soil data

@@

  ## crop to reduce memory usage
  box = c(124.5, 127.5, -9.2, -8.3) # Timor-Leste
  rst <- terra::rast(destfile)
  rst <- terra::crop(rst, terra::ext(box))
  writeRaster(rst, filename = destfile, overwrite = TRUE) 
}

@@

The research was carried out in the center of the Baucau district, on the north-eastern coast of Timor-Leste. Two communities, or suco (group of villages), were selected to represent two contrasting topo-geographical, climatic and historical cases: (i) Gariuai suco (long-lat coordinates: `r site_summary[site=="Gariuai", coord]`) is located on a plateau of Baucau limestone, with an annual rainfall of `r site_summary[site=="Gariuai", prec]` mm and an altitude ranging 300-750 m; and (ii) the village of Osso Luga in Samalari suco (long-lat coordinates: `r site_summary[site=="Osso Luga", coord]`) is located on the Matebian foothill based on Bobonaro Scaly Clay, with an annual rainfall of `r site_summary[site=="Osso Luga", prec]` mm and an altitude ranging `r site_summary[site=="Osso Luga", elev]` m (Figure\ \@ref(fig:map-sites)). The average soil properties differed between sites (Table\ S\@ref(tab:site-soil)), with Gariuai having Clay Loam soil texture compared to Osso Luga having clay soil texture. The two sites also have had different histories. Gariuai has been inhabited for at least 50 years, while Osso Luga was abandoned during the Indonesian occupation, and was only repopulated after 2003. During the period of abandonment, the various agroforestry systems in Osso Luga lay fallow.

@@

We used the typology described in @Cogne2024, developed in the same study area. This typology was based on the one proposed by @Nair1993;@Nair2021, and adapted to the local context with semi-structured interviews conducted in the same study area [@Cogne2024]. The five AFS identified were: crop system including a wooded fallow phase (CF), Silvopastoral (SP), Young agroforest (YA), Home garden (HG) and Forest garden (FG). CF systems are plantations of a few non-perennial crops (such as maize, sweet potatoes, groundnuts and squash) which are left fallow after 1 to 3 years. SP land is a vast area (generally 200 to 500 ha) where cattle are grazed under a layer of trees. YAs are agricultural plots where the farmer has been planting rows of trees, spaced at around 5x10m, for two to 10 years, usually with the help of projects. HGs are plots of land close to houses, where crops and livestock (poultry, pigs, cattle, etc.) have been grown in the shade of numerous trees and palms for multiple uses (wood, fruit, fibre, etc.) for more than five years. FGs are generally former HGs (over 50 years old), which have been abandoned, for example as a result of the war, and are now used mainly for gathering with limited human influence.

@@

Taxonomic identification was carried out using tetum vernacular names, which were then linked to scientific names using the iNaturalist and Pl\@ntNet applications, Geoffrey Hull’s monograph [@Hull2006], the “Useful tropical plants” database [@Fern2014] as well as the Australian Northern Territory government's Native Plants Herbarium [@NT2013]. The correspondence between tetum names and scientific names is provided in the supplemental material (Table S\@ref(tab:tetum-names)).

@@

# stack tree tables
data_tree <- rbind(data_large[, col_keep], data_small[, col_keep])

# merge with plot information
data_tree <- merge(data_tree, data_plot, by = "N_PLOT")

# convert data_tree into data.table 
setDT(data_tree)

@@

# remove NA or <30 cm circumference values or < 100 cm outside subplots
data_tree <- subset(data_tree, 
                   !is.na(DIAM_130_CM) & 
                     DIAM_130_CM >= 30/pi &
                     (DIAM_130_CM >= 100/pi | grepl("Q", N_Q_ID)))

# split scientific names into genus and species
names <- tstrsplit(data_tree$SCIENTIFIC_NAME, " ")

# get genera
data_tree$GENUS <- names[[1]]

# get species
data_tree$SPECIES <- names[[2]]

# Timoneus should be Timonius
data_tree$GENUS[!is.na(data_tree$GENUS) & data_tree$GENUS == "Timoneus"] <- "Timonius"

# remove one tree with DBH = 0
data_tree <- subset(data_tree, DIAM_130_CM > 0)

@@

# get wood density
data_wd <- getWoodDensity(genus = data_tree$GENUS, species = data_tree$SPECIES)

# estimate mean AGC
data_tree$mean <- computeAGB(
  D = data_tree$DIAM_130_CM, 
  WD = data_wd$meanWD, 
  H = data_tree$HEIGHT
) * 0.4713 # conversion factor from biomass to carbon

@@

data_sim_agc <- data.table(
  data_tree,
  AGBmonteCarlo(
    D = data_tree$DIAM_130_CM, 
    WD = data_wd$meanWD, 
    errWD = data_wd$sdWD, 
    H = data_tree$HEIGHT, 
    errH = data_tree$HEIGHT*cvH, 
    Dpropag = "chave2004", 
    n = 1000,
    Carbon = TRUE
  )$AGC_simu
) |> 
  melt(
    measure.vars = c("mean", paste0("V", 1:1000)),
    variable.name = "iter",
    value.name =  "AGC"
  )

@@

## sum mean biomass in quadrats and entire plot
data_sim[, value := value + valueSmall]

## add zeros for plots with no tree above 10 cm DBH
# deal with non detectable values: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC9216349/
# use uniform distribution between zero and limit of detectability
data_sim_empty = expand.grid(N_PLOT = setdiff(data_plot$N_PLOT, data_sim$N_PLOT), 
                             iter = unique(data_sim$iter), 
                             variable = c("AGC", "BGC"))|> 
  setDT() |> merge(data_plot[, c("N_PLOT", "ID_VILLAGE", "ID_AF")]) 
data_sim_empty$value = 0

# limit of detectability: tree of 30 cm CBH, mean WD and height (for this DBH)
# LOD = computeAGB(D = 30/pi, 
#                  WD = unique(data_wd$meanWD[data_wd$genus=="NR"]), 
#                  H = data_tree[DIAM_130_CM == 30/pi, mean(HEIGHT)])
# data_sim_empty[iter != "mean", AGC := runif(sum(iter != "mean"), 0, LOD)]
# data_sim_empty[iter == "mean", AGC := LOD/2]
# RS = mean(data_sim_agc$RS[data_sim_agc$DIAM_130_CM==30/pi])
# data_sim_empty[, BGC := AGC * RS]

# data_sim_empty = melt(data_sim_empty, measure.vars = c("AGC", "BGC"))

data_sim = rbind(data_sim[, colnames(data_sim_empty), with = FALSE], data_sim_empty)
```

@@

Soil organic carbon stocks (in Mg\ C\ ha$^{-1}$) were then calculated using the method recommended by @Poeplau2017 (M4), which correctly takes into account the coarse fraction of the soil:

@@

# add total carbon stocks
data_sim = rbind(data_sim, data_sim[, .(value = sum(value[variable != "SOC10"]), variable = "TOT"), .(N_PLOT, site, ID_AF, iter)])
# group total C stocks in plots 6 and 8 (otherwise will underestimate CF total C
# stocks); this will not be used for analysis, only to display average C stocks
# per AFS
# data_sim <- rbind(
#   data_sim[variable != "TOT" | !N_PLOT %in% c(6, 8)], 
#   data_sim[variable == "TOT" & N_PLOT %in% c(6, 8), .(value = sum(value), N_PLOT = 6), .(site, ID_AF, variable, iter)]
# )
```

@@

The estimated AGC and SOC stocks were log-transformed for all statistical tests, because we expect them to be log-normally distributed as they only take positive values. We removed one CF plot at Gariuai from the following statistical tests, as it did not contain any trees exceeding the minimum measurement size (CBH = 30 cm) and its AGC stocks were therefore estimated at 0\ Mg\ C\ ha$^{-1}$ (even if there were trees < 30 cm CBH) and could not be log-transformed. After testing for normality (Shapiro-Wilks test) and homoscedasticity (Levene’s test), we performed a two-way ANOVA on the log-transformed estimates of AGC and SOC stocks, grouped by AFS and site. To test the relationship between log-transformed SOC and AGC stocks, we fitted two linear mixed models. We predicted SOC stocks at two different depths (either 0-10 or 0-30 cm) as a function of AGC stocks (fixed effect) and site (Samalari or Osso-Luga; random effect).

@@

```{r group-comparison}
# shapiro wilk test - normality
dataC_plot[variable %in% c("AGC", "SOC") & plot != 6, .(
  shap = shapiro.test(log(mean))$p.value
), .(afs, variable)]
# levene test of homoscedasticity
dataC_plot[variable %in% c("AGC", "SOC") & plot != 6, 
           leveneTest(log(mean) ~ afs*site)$`Pr(>F)`[1], .(variable)]

# 2-way anova 
anova_table = 
  dataC_plot[variable %in% c("AGC", "SOC") & plot != 6, 
             .(pvalue = round(anova(lm(log(mean) ~ site + afs))$`Pr(>F)`[1:2], 3)), 
             .(variable)]
anova_table$factor = rep(c("site", "afs"), 2)
```

```{r lmereg}
data_reg = melt(dataC_plot[plot != 6], measure.vars = c("mean", "lwr", "upr"), variable.name = "stat") |> 
  dcast(plot + site+ afs ~ variable + stat)

@@

Total carbon stocks varied between `r dataC_plot[variable=="TOT", round(min(mean))]` and `r dataC_plot[variable=="TOT", round(max(mean))]`\ Mg\ C\ ha$^{-1}$, with an average of `r dataC_plot[variable=="TOT", round(mean(mean))]`\ Mg\ C\ ha$^{-1}$ and a coefficient of variation of `r dataC_plot[variable=="TOT", round(sd(mean)/mean(mean)*100)]`%  (Table \@ref(tab:summary); Figure\ \@ref(fig:barplot)). The AFS with the lowest total carbon stocks was the `r dataC_afs[variable=="TOT", afs[which.min(mean)]]` with an average of `r dataC_afs[variable=="TOT", round(mean[which.min(mean)])]`\ Mg\ C\ ha$^{-1}$; the AFS with the highest total carbon stocks was the `r dataC_afs[variable=="TOT", afs[which.max(mean)]]` with an average of `r dataC_afs[variable=="TOT", round(mean[which.max(mean)])]`\ Mg\ C\ ha$^{-1}$ (Table\ \@ref(tab:summary); Figure\ \@ref(fig:barplot)).

@@

AGC stocks varied between `r dataC_plot[variable=="AGC", round(min(mean))]` and `r dataC_plot[variable=="AGC", round(max(mean))]`\ Mg\ C\ ha$^{-1}$, with an average of `r dataC_plot[variable=="AGC", round(mean(mean))]`\ Mg\ C\ ha$^{-1}$ and a coefficient of variation of `r dataC_plot[variable=="AGC", round(sd(mean)/mean(mean)*100)]`%  (Table \@ref(tab:summary); Figure\ \@ref(fig:barplot)). AGC stocks varied significantly between AFS (ANOVA p-value: `r subset(anova_table, variable=="AGC" & factor == "afs")$pvalue`, Table \@ref(tab:anova)), but not between site (ANOVA p-value = `r subset(anova_table, variable=="AGC" & factor == "site")$pvalue`, Table \@ref(tab:anova)). The AFS with the lowest AGC stocks was the `r dataC_afs[variable=="AGC", afs[which.min(mean)]]` with an average of `r dataC_afs[variable=="AGC", round(mean[which.min(mean)])]`\ Mg\ C\ ha$^{-1}$; the AFS with the highest AGC stocks was the `r dataC_afs[variable=="AGC", afs[which.max(mean)]]` with an average of `r dataC_afs[variable=="AGC", round(mean[which.max(mean)])]`\ Mg\ C\ ha$^{-1}$ (Table\ \@ref(tab:summary); Figure\ \@ref(fig:barplot)).

@@

SOC stocks varied between `r dataC_plot[variable=="SOC", round(min(mean))]` and `r dataC_plot[variable=="SOC", round(max(mean))]`\ Mg\ C\ ha$^{-1}$, with an average of `r dataC_plot[variable=="SOC", round(mean(mean))]`\ Mg\ C\ ha$^{-1}$ and a coefficient of variation of `r dataC_plot[variable=="SOC", round(sd(mean)/mean(mean)*100)]`%  (Table \@ref(tab:summary); Figure\ \@ref(fig:barplot)). No significant differences in SOC were observed between the five AFS studied (ANOVA p-value: `r subset(anova_table, variable=="SOC" & factor == "afs")$pvalue`, Table \@ref(tab:anova)). However, the measured SOC stocks differed between sites (ANOVA p-value: `r subset(anova_table, variable=="SOC" & factor == "site")$pvalue`, Table \@ref(tab:anova)). The SOC stocks in Gariuai were on average higher (`r dataC_site[site=="Gariuai" & variable=="SOC", round(mean)]`\ Mg\ C\ ha$^{-1}$) than those in Osso Luga (`r dataC_site[site=="Osso Luga" & variable=="SOC", round(mean)]`\ Mg\ C\ ha$^{-1}$; Figure\ \@ref(fig:barplot)).

@@

## Large C stocks in traditional AFS in Timor-Leste

Compared to values reported in the literature for other AFS, the carbon stocks in this study were particularly high. The value of SOC stocks was similar to other reported values [@Feliciano2018], but some of the AFS studied here have particularly high stocks in their biomass [compared to the average of 74\ Mg\ C\ ha$^{-1}$ in @Ma2020]. The AFS with the highest carbon stocks, forest gardens, had stocks close to those of old-growth tropical forests [270\ Mg\ C\ ha$^{-1}$; @Lal2005]. These high carbon stocks could be explained by the fact that forest gardens have generally grown for more than 50 years, with little human intervention (Table S1). In terms of structure and carbon stocks, forest gardens should therefore resemble old secondary forests, which have been shown to recover more than 60% of the AGC stocks and almost all the SOC stocks of an old forest [@Poorter2021]. In addition, FGs are generally located close to villages, where soils are enriched by organic waste and animal manure, which can enhance tree growth [@Madejon2016].

@@

The location of AFS in the landscape is not random and could also explain the observed differences in AGC stocks. For example, most HGs are located close to houses, in alluvial valleys that are well supplied with water and where the soil is enriched with organic waste and animal manure, which is likely to increase soil fertility [@Palm2001]. The same applies to FGs, which are former HGs that have been more or less abandoned, and YAs, where the farmer has chosen the site with the intention of turning it into a HG. We therefore expect HGs, FGs and YAs to be located in areas that are more favorable for tree growth [@Wagner2012;@Madejon2016;@Treuer2018]. Conversely, CF and SP fields are mainly located on limestone plateaus where water is deep and soils are shallow, which could reduce tree growth and biomass [@Imada2008].

@@

## How to conserve and improve carbon stocks in traditional AFS in Timor-Leste?

Although agroforestry can make agriculture more resilient while improving its carbon footprint and biodiversity conservation value [@Cardinael2021], poor adaptation to the local context can explain many failures to adopt new agroforestry techniques [@Coe2014]. Conversely, some traditional agroforestry techniques have evolved in a particular context (biophysical, social and economic) and are therefore adapted to it [@Gouyon1993;@Aumeeruddy1994;@Peltier1996]. This is the case of the traditional AFS studied here: their heritage and socio-economic values [as shown by @Cogne2024], in addition to their high carbon stock value, confirm the importance of their conservation, both nationally and internationally.

@@

# Authors' contribution

All authors contributed to the design of the sampling protocol. MC carried out the fieldwork with the help of RP, AT and VF [mention students?]. CP analyzed the data with the help of VF and MC. CP prepared the manuscript and all authors contributed to the drafting and revision of the manuscript.

# Acknowledgements

# Data availability

@@

```{r tab.id = "summary", results = "show", tab.cap = "Mean and associated 95% confidence intervals (between parentheses) of total, aboveground carbon (AGC), belowground carbon (BGC), and soil organic carbon (SOC) stocks in each agroforestry system."}
# calculate median and 95% confidence interval at site level
dataC_afs[, value := paste0(
  signif(mean, 3), " (",
  signif(lwr_mean, 3), "-", 
  signif(upr_mean, 3), ")")] |> 
  subset(variable != "SOC10") |> 
  dcast(afs ~ variable, value.var = "value") |> 
  flextable(col_keys = c("afs", "TOT", "AGC", "BGC", "SOC"), 
            cwidth = c(0.9, 1.2, 1.3, 1.3, 1.2)) |> 
  compose(i = 1, j = "afs", part = "header", value = as_paragraph("Agroforestry system")) |>
  compose(i = 1, j = "TOT", part = "header", value = as_paragraph( "Total C stocks\n(Mg C ha", as_sup("-1"), ")"))|> 
  compose(i = 1, j = "AGC", part = "header", value = as_paragraph( "AGC stocks\n(Mg C ha", as_sup("-1"), ")"))|> 
  compose(i = 1, j = "BGC", part = "header", value = as_paragraph( "BGC stocks\n(Mg C ha", as_sup("-1"), ")"))|> 
  compose(i = 1, j = "SOC", part = "header", value = as_paragraph( "SOC stocks\n(Mg C ha", as_sup("-1"), ")"))
```
